Does It Scale? Symbolic Execution of PHP Web Applications

### 2017-01-22 paper.tex
updated table

Based on observations we identify conceptual limitations of symbolic
execution for PHP.  Dynamic features such as indirect function calls and
include expression require concrete information to be evaluated properly. Since
these expressions may be assembled at runtime and can contain symbolic
information, for many cases this restricts a symbolic execution engine from
further execution.
Although some effort to statically approximate include expression can be spent,
it is often non-trivial since expressions can contain information coming from
various inputs including databases, user inputs or configuration files. We
empirically evaluate our observations for a corpus including large, actively maintained PHP systems.

@@

For our implementation of a symbolic execution engine for PHP, Oak, we have
chosen a test-driven approach with continuous integration using regression
tests based on \sf{WordPress} and \sf{SchoolMate} (see Section
\ref{sec:experiment_setup}). For \sf{SchoolMate}, a rather straightforward and
simple system, which we chose to compare our approximation results to those of
\sf{PhpSync}, we could achieve similar output approximation results. For
\sf{WordPress} though, we did not achieve high code coverage or an accurate
approximation for \sf{WordPress} as we faced a number of obstacles hindering
symbolic execution in different ways: The following Subsection
\ref{sec:experience_report} documents our design choices we made, apparent trade-offs we recognized, and limitations
we faced on our way towards a scalable symbolic execution engine; and what we
have learned from bypassing limitations faced in spite of manual effort for
future work. Subsection \ref{sec:experience_results} compares our observations
and results to \emph{PhpSync} and presents the motivation for our empirical evaluation of our
tool symbolic execution engine in Section \ref{sec:evaluation}.

@@

Second, function can either be called directly by statically providing the name
of the function (or method), or indirectly. Indirect function calls in PHP are
enabled by using built-in functions like \tt{call\_user\_func},
\tt{call\_user\_func\_array} or simply passing a string value with a direct
function call to the \tt{eval} function. The first two functions take as
arguments the name of the function two call, and all arguments to that
functions either as additional arguments or a single associative array
respectively.
The \tt{eval} function evaluates a given PHP expression passed as a string
literal.
Again, we encountered issues similar to dynamic include expressions with
indirect function call mechanisms: The example in Figure
\ref{fig:indirect_calls} illustrates two functions \tt{header\_serif} and
\tt{header\_sserif} which both print a HTML headline. Depending on whether the
value of the variable \tt{\$style} retrieved from a database is \tt{serif} or
\tt{sserif}, a different function is called. Given that the function name is 
concrete at runtime, the desired function can be called. Otherwise, if the
function name is assembled at runtime and contains symbolic values, no function
can be called as the name is ambiguous or unknown. For this example, symbolic
execution is not able to determine which function to call indirectly.

@@

if (...) {
   $person = new Person("Alice");
} else {
   $person = new Person("Bob");
}
$person->greet("Good Morning, ");
\end{lstlisting}
	\caption{
		Dynamic dispatch: Multiple targets for method calls
		\label{fig:multi_target}
	}
	\end{subfigure}
	\vspace{2mm}
	\hfill
	\begin{subfigure}[b]{0.48\textwidth}

@@

require_once TEMPLATE . $template;
\end{lstlisting}
	\caption{
		Dynamic Include Resolution
		\label{fig:dynamic_includes}
	}
	\end{subfigure}
	\vspace{2mm}
	\hfill
	\begin{subfigure}[b]{0.48\textwidth}

@@

$font_style = mysql_result(...);
call_user_func('header_' . $font_style, 'Title');
\end{lstlisting}
	\caption{
		Function calls by indirect invocation mechanisms
		\label{fig:indirect_calls}
	}

	\end{subfigure}
	\caption{
		Code Examples of defective code features
		\label{fig:defective_code_features}
	}
\end{figure}

@@

\begin{table*}[t]
\centering 
	\begin{tabular}{lrp{4cm}rrrr}
	\toprule
	\textbf{System} & \textbf{Version} & \textbf{Classification} & \textbf{SLOC} &
	\textbf{\#files} & \textbf{\#OCs} & \textbf{\#includes}\\
	\midrule
	AddressBook & 8.2.5.2 &  & 51,907 & 239  & 1009 & 186\\
	SchoolMate & 1.5.4 &  & 8,118 & 65  & 853 & 88\\
	TimeClock & 1.04 &  & 20,800 & 63  & 7920 & 306\\
	WebChess & 1.0.0 &  & 5,219 & 28  & 470 & 56\\
	\midrule
	Drupal & 7.5.0 & CMS & 52,464 & 125  & 3569 & 749\\
	phpBB & 3.1.9 &  & 327,371 & 1,398  & 3606 & 206\\
	phpMyAdmin & 4.6.3 &  & 303,582 & 871  & 7103 & 571\\
	\midrule
	Anchor & 0.12.1 &  & 15,054 & 201 & 987 & 32\\
	Kirby & XXX & CMS & XXX & XXX  & 654 & 23\\
	Automad & XXX & CMS & XXX & XXX  & 655 & 8\\
	Monstra & XXX & CMS & XXX & XXX  & 1934 & 48\\
	Nibbleblog & XXX & CMS & XXX & XXX  & 1013 & 28\\
	\bottomrule
	\end{tabular}
	\caption{Corpus of twelve PHP systems. The file count includes files with a .php,
	.inc, .bit or .module extension.}
	\label{corpus}
\end{table*}

@@

Given a symbolic value, indirect call mechanisms like callback-commands are
likely to fail since target function name and the symbolic value do not match.
This also applies for the evaluation of include expressions as concrete values
representing include targets can be included, but any include target containing
symbolic values is ambiguous. Since it is unlikely that all functions containing
missed output candidates are dead code having no call site (direct or
indirect) at all, in Section \ref{WhyDidFunctionCallsFail} we further
investigate how many functions fail due to inaccurately resolved indirect
function calls. Moreover, we evaluate the accuracy of evaluation of include
expressions in the next section.

@@

Moreover we need to take into account that one missed function execution can
result in missing even more function calls (direct or indirect) and executions.
Thus, for functions containing not-reached output candidates we measure whether
they can be (1) only, (2) partially or (3) never accessed transitively from
functions that have no direct call sites. We refer to those functions (or
access points) having no direct call site as callback candidates since the only
way to reach them is through indirect mechanisms. Measuring how many
not-reached output candidates can be explained by functions that are callback
candidates helps to understand whether, and if so, indirect call mechanisms and
their usage may impact our analysis’ code coverage.

@@

For failed function calls to functions containing not-reached output
candidates we measured how many of those functions were only, or
partially accessibly through indirect calling mechanisms (see Section
\ref{sec:experiment_dynamicfeatures}). As Figure
\ref{fig:output_candidate_explanation} illustrates, around 80 percent (except
for \sf{TimeClock}) all missed output candidates of each system were only
accessible through callback candidates, i.e., functions that are only
accessible through indirect calling mechanisms.

@@

\paragraph{Dynamic features are problematic.}
From out observations with \sf{WordPress} in Section \ref{sec:3} we learn that
dynamic language features, in particular include expressions and indirect function
calls, in combination with symbolic values cannot be evaluated accurately since
concrete values are required. For our observations we were able to explain most
of the missed output candidates with indirect function calls. For include
expressions we are indeed able to resolve most expressions successfully, yet we
do not distinguish between static and dynamic includes in this case study as
this would exceed the scope of our paper. However, we believe it is plausible
to assume that defective include expressions can cause a cascade of missing
files and subsequent includes.
Throughout the experiment we have seen that dynamic features, are prevalent for
modern PHP systems as they serve a genuine purpose in the programming language.
Nevertheless their usage in situations when combined with assumed symbolic
values restricts symbolic execution and any tool leveraging results obtained
from that.

@@

%Tools using symbolic execution for static output approximation \cite{artzi_finding_2008,artzi_finding_2010}
%\textcolor{blue}{\blindtext}

>>>>

Based on observations we identify conceptual limitations of symbolic
execution for PHP.  Dynamic features such as indirect function invocations and
include expression require concrete information to be evaluated properly. Since
these expressions may be assembled at runtime and can contain symbolic
information, for many cases this restricts a symbolic execution engine from
further execution.
Although some effort to statically approximate include expression can be spent,
it is often non-trivial since expressions can contain information coming from
various inputs including databases, user inputs or configuration files. We
empirically evaluate our observations for a corpus including large, actively maintained PHP systems.

@@

For our implementation of a symbolic execution engine for PHP, Oak, we have
chosen a test-driven approach with continuous integration using regression
tests based on \sf{WordPress} and \sf{SchoolMate} (see Section
\ref{sec:experiment_setup}). For \sf{SchoolMate}, a rather straightforward and
simple system, which we chose to compare our approximation results to those of
\sf{PhpSync}, we could achieve similar output approximation results. For
\sf{WordPress} though, we did not achieve high code coverage or an accurate
approximation for \sf{WordPress} as we faced a number of obstacles hindering
symbolic execution in different ways: The following Subsection
\ref{sec:experience_report} documents our design choices we made, apparent trade-offs we recognized, and limitations
we faced on our way towards a scalable symbolic execution engine; and what we
have learned from bypassing limitations faced in spite of manual effort for
future work. Subsection \ref{sec:experience_results} compares our observations
and results to \sf{PhpSync} and presents the motivation for our empirical
evaluation of our tool symbolic execution engine in Section \ref{sec:evaluation}.

@@

Second, function can either be invocated directly by statically providing the
name of the function (or method), or indirectly. Indirect, or dynamic function
invocation in PHP are enabled by using built-in functions like \tt{call\_user\_func},
\tt{call\_user\_func\_array} or simply passing a string value with a direct
function call to the \tt{eval} function. The first two functions take as
arguments the name of the function two call, and all arguments to that
functions either as additional arguments or a single associative array
respectively.
The \tt{eval} function evaluates a given PHP expression passed as a string
literal.
Again, we encountered issues similar to dynamic include expressions with
\ref{fig:indirect_calls} illustrates two functions \tt{header\_serif} and
\tt{header\_sserif} which both print a HTML headline. Depending on whether the
value of the variable \tt{\$style} retrieved from a database is \tt{serif} or
\tt{sserif}, a different function is called. Given that the function name is 
concrete at runtime, the desired function can be called. Otherwise, if the
function name is assembled at runtime and contains symbolic values, no function
can be called as the name is ambiguous or unknown. For this example, symbolic
execution is not able to determine which function to call indirectly.

@@

if (...) {
   $person = new Person("Alice");
} else {
   $person = new Person("Bob");
}
$person->greet("Good Morning, ");
\end{lstlisting}
	\caption{
		Dynamic Dispatch : Multiple target objects for method invocation \tt{greet()} 
		\label{fig:multi_target}
	}
	\end{subfigure}
	\vspace{2mm}
	\hfill
	\begin{subfigure}[b]{0.48\textwidth}

@@

require_once TEMPLATE . $template;
\end{lstlisting}
	\caption{
		Dynamic Include Resolution: The value of \tt{\$template} is retrieved from a
		database query result.
		\label{fig:dynamic_includes}
	}
	\end{subfigure}
	\vspace{2mm}
	\hfill
	\begin{subfigure}[b]{0.48\textwidth}

@@

$font_style = mysql_result(...);
call_user_func('header_' . $font_style, 'Title');
\end{lstlisting}
	\caption{
		Dynamic Function Invocation using \tt{call\_user\_func}
		\label{fig:indirect_calls}
	}

	\end{subfigure}
	\caption{
		Code examples for dynamic features
		\label{fig:defective_code_features}
	}
\end{figure}

@@

% count lines
%find . -regextype "posix-extended" -iregex '.*\.(php|bit|inc|module)' | xargs wc -l

% count files
%find . -regextype "posix-extended" -iregex '.*\.(php|bit|inc|module)' | wc -l

\begin{table*}[t]
\centering 
	\begin{tabular}{lrp{5cm}rrrr}
	\toprule
	\textbf{System} & \textbf{Version} & \textbf{Classification} & \textbf{SLOC} &
	\textbf{\#files} & \textbf{\#Output candidates} & \textbf{\#Include
	Expressions}\\
	\midrule
	AddressBook & 8.2.5.2 & Web-based address and phone book & 51,907 & 239  & 1009
	& 186\\
	SchoolMate & 1.5.4 & School administration & 8,118 & 65  & 853 & 88\\
	TimeClock & 1.0.4 & Web-based time tracking & 20,800 & 63  & 7920 & 306\\
	WebChess & 1.0.0 & Chess server software & 5,219 & 28  & 470 & 56\\
	\midrule
	Drupal & 7.5.0 & Content Management System &  52,464 & 125  & 3569 & 749\\
	phpBB & 3.1.9 & Forum & 327,371 & 1,398  & 3606 & 206\\
	phpMyAdmin & 4.6.3 & Web-based MySQL administration & 303,582 & 871  & 7103 &
	571\\
	\midrule
	Anchor & 0.12.1 & Content Management System & 15,054 & 201 & 987 & 32\\
	Kirby & 2.4.1 & Content Management System & 51.778 & 435  & 654 & 23\\
	Automad & 0.10 & Content Management System & 13.978 & 64  & 655 & 8\\
	Monstra & 3.0.4 & Content Management System & 41.102 & 490  & 1934 & 48\\
	Nibbleblog & 4.0.5 & Content Management System & 20.376 & 328  & 1013 & 28\\
	\bottomrule
	\end{tabular}
	\caption{Corpus of twelve PHP systems. The file count includes files with a .php,
	.inc, .bit or .module extension.}
	\label{corpus}
\end{table*}

@@

Given a symbolic value, indirect function invocation mechanisms like
callback-commands are likely to fail since target function name and the symbolic value do not match.
This also applies for the evaluation of include expressions as concrete values
representing include targets can be included, but any include target containing
symbolic values is ambiguous. Since it is unlikely that all functions containing
missed output candidates are dead code having no call site (direct or
indirect) at all, in Section \ref{WhyDidFunctionCallsFail} we further
investigate how many functions fail due to inaccurately resolved indirect
function invocations. Moreover, we evaluate the accuracy of evaluation of
include expressions in the next section.

@@

Moreover we need to take into account that one missed function execution can
result in missing even more function calls (direct or indirect) and executions.
Thus, for functions containing not-reached output candidates we measure whether
they can be (1) only, (2) partially or (3) never accessed transitively from
functions that have no direct call sites. We refer to those functions (or
access points) having no direct call site as callback candidates since the only
way to reach them is through dynamic function invocation mechanisms. Measuring
how many not-reached output candidates can be explained by functions that are callback
candidates helps to understand whether, and if so, indirect function invocation
mechanisms and their usage may impact our analysis’ code coverage.

@@

For failed function calls to functions containing not-reached output
candidates we measured how many of those functions were only, or
partially accessibly through indirect function invocation mechanisms (see
Section \ref{sec:experiment_dynamicfeatures}). As Figure
\ref{fig:output_candidate_explanation} illustrates, around 80 percent (except
for \sf{TimeClock}) all missed output candidates of each system were only
accessible through callback candidates, i.e., functions that are only
accessible through indirect calling mechanisms.

@@

\paragraph{Dynamic features are problematic.}
From out observations with \sf{WordPress} in Section \ref{sec:3} we learn that
dynamic language features, in particular include expressions and indirect function
invocations, in combination with symbolic values cannot be evaluated accurately
since concrete values are required. For our observations we were able to explain most
of the missed output candidates with indirect function invocations. For include
expressions we are indeed able to resolve most expressions successfully, yet we
do not distinguish between static and dynamic includes in this case study as
this would exceed the scope of our paper. However, we believe it is plausible
to assume that defective include expressions can cause a cascade of missing
files and subsequent includes.
Throughout the experiment we have seen that dynamic features, are prevalent for
modern PHP systems as they serve a genuine purpose in the programming language.
Nevertheless their usage in situations when combined with assumed symbolic
values restricts symbolic execution and any tool leveraging results obtained
from that.

@@
